---
title: "CoordCleaning_MasterDF"
author: "emlombardi"
date: "2022-10-02"
output: html_document
---

#Objective: clean up coordinates in the master dataframe The csv file loaded here will be the occurrences for all plants from NM. It is created in the `Occ data wrangling` script.

```{r libraries and data}
#load libraries
library(tidyverse)
library(rgbif)
library(sp)
library(countrycode)
library(CoordinateCleaner)


#load RDA file instead
load(file="script1_output.RData")
load(file="rawDB_TNRSharm_output.RData") #If trying different harmonization methods; this is for TNRS alone

```

#Visualize the coordinates by mapping This helps make sure things are lining up in space

```{r mapping}
world.inp  <- map_data("world")

ggplot()+
  geom_map(data=world.inp, 
           map=world.inp, 
           aes(x=long, y=lat, map_id=region), 
           fill = "grey80")+
  xlim(min(dat.df$decimalLongitude, na.rm = T), 
       max(dat.df$decimalLongitude, na.rm = T))+
  ylim(min(dat.df$decimalLatitude, na.rm = T), 
       max(dat.df$decimalLatitude, na.rm = T))+
  geom_point(data = dat.df, 
             aes(x = decimalLongitude, y = decimalLatitude, color=source),
             size = 1)+
  coord_fixed()+
  theme_bw()+
  theme(axis.title = element_blank())


```

#Clean coordinates based on metadata 
There might be a few ways to tidy things up based on coordinate uncertainty, state locality, etc.

Note that I'm going to rename `dat.df` to `dat_cl`, which will be the name of the main dataframe as we pass it through cleaning scripts. 

```{r coord cleaning}
# remove records without coordinates 
dat_cl <- dat.df %>% filter(!is.na(decimalLongitude)) %>% filter(!is.na(decimalLatitude)) ###NONE removed. They all have coordinates (because that's how we built this dataset...so this is a good thing)
dim(dat_cl) #264,744 records (feb 2023)

# remove records with low coordinate precision
hist(dat_cl$coordinateUncertaintyInMeters/1000, breaks = 100)

dat_cl <- dat_cl %>% filter(coordinateUncertaintyInMeters/1000 <= 100 | is.na(coordinateUncertaintyInMeters)) #Removed 285 records (Feb2023)

# remove unsuitable data sources, especially fossils
table(dat_cl$basisOfRecord) #Why are there any human observations involved???? It just happens. Messes and stuff.

dat_cl<-dat_cl%>% filter(!basisOfRecord=="HumanObservation", !basisOfRecord=="Observation", !basisOfRecord=="MachineObservation", !basisOfRecord=="LivingSpecimen") #this removes the strange 1163 records (feb2023)

#remove impossible coordinates
dat_cl<-dat_cl%>%filter(!decimalLatitude>90, !decimalLatitude< -90) #removes impossible records. There are none now (feb2023)

```


#Clean coordinates with CoordinateCleaner package

```{r cleaning package}
library(CoordinateCleaner)
dat_cl<-data.frame(dat_cl)

flags <- clean_coordinates(x = dat_cl, lon = "decimalLongitude", lat = "decimalLatitude", species = "Accepted_species", tests = c("capitals", "centroids", "equal", "gbif", "zeros", "seas"), seas_ref = buffland)  # flagged zero records! Our cleaning is working (feb2023)

```

#Map to visualize difference between cleaned and uncleaned coords

```{r map differences}

# first plot showing the retain records in green and the removed records in red. We don't have any flags, so this isn't super interesting.
ggplot()+
  geom_map(data=world.inp, 
           map=world.inp, 
           aes(x=long, y=lat, map_id=region), fill = "grey80")+
  xlim(min(dat.df$decimalLongitude, na.rm = T),
       max(dat.df$decimalLongitude, na.rm = T))+
  ylim(min(dat.df$decimalLatitude, na.rm = T), 
       max(dat.df$decimalLatitude, na.rm = T))+
  geom_point(data = dat.df, 
             aes(x = decimalLongitude, y = decimalLatitude),
             colour = "darkred", size = 1)+
  geom_point(data = dat_cl, 
             aes(x = decimalLongitude, y = decimalLatitude),
             colour = "lightpink3", size = .2, alpha=0.5)+
  coord_fixed()+
  theme_bw()+
  theme(axis.title = element_blank())

# second plot showing only the retained species
ggplot()+
  geom_map(data=world.inp, map=world.inp,
           aes(x=long, y=lat, map_id=region), 
           fill = "grey80")+
  xlim(min(dat_cl$decimalLongitude, na.rm = T), 
       max(dat_cl$decimalLongitude, na.rm = T))+
  ylim(min(dat_cl$decimalLatitude, na.rm = T), 
       max(dat_cl$decimalLatitude, na.rm = T))+
  geom_point(data = dat_cl, 
             aes(x = decimalLongitude, y = decimalLatitude),
             size = .5, alpha=.2)+
  coord_fixed()+
  theme_bw()+
  theme(axis.title = element_blank())


#IDEA: can I also map or grab the coordinates that were cleaned out of the dataset? It should just be the observations, I think, though

```

#Save the cleaned data
This should be the data useful for all sorts of analyses (richness, temporal, etc). 

```{r save}
write_csv(dat_cl, "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Occurrence Data/dat_cl_Feb2023_script2.csv")

#save full working environment
save(dat_cl, flags, file="script2_output.RData")

```
This script removes 1,448 coordinates based on coordinate uncertainty and general spatial flags