---
title: "4. Download environmental data layers"
author: "emlombardi"
date: "2022-12-02"
output: html_document
---
#Summary
This code is a stand-alone script that will allow you to download enviornmental data layers from open sources. In this script we will also standardize the resolution of all rasters, the projection (specific???) and the extent (we clip to New Mexico in the southwestern United State). These are necessary steps before individual layers are stacked into a RasterBrick (??). 

Useful information about accessible remotely sensed data sources: https://rspatial.org/rs/1-introduction.html

Inputs:
Outputs:

```{r libraries}
library(maps)
library(mapview)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(spatstat)
library(stars)
library(stringr)
library(tidyverse)
library(spdep)
library(splm)
library(terra) #unloaded issue
library(geodata) #unloaded issue
library(ggplot2)
library(ggpubr)
library(prism)
library(AOI)
library(climateR)
library(raster)
library(rasterVis)

theme_set(theme_pubr())


#Area of interest. ADD THIS TO THE FIRST WKT DEFINITION
aoi=aoi_get(state="NM") %>% st_transform(crs=4326)
class(aoi)
plot(aoi$geometry) #sf boundary of NM state


```

#Load data

```{r}
##data from last cleaning step
load(file="GitHub_script3_georefCleaned.RData")


#create spatial feature from coordinates 

c2 <- as.data.frame(dat_sf) %>% 
  st_as_sf(coords=c("decimalLongitude","decimalLatitude"), crs=4326, remove=FALSE) 

class(coords)
class(c2) #sf format


````

###Try lining coords up with NM

```{r}
###Load shapefile of study area (in our case, New Mexico)
bb.nm<-tigris::tracts(state = 'NM')
bb.nm<-st_as_sf(bb.nm) %>% st_transform(crs=4326)

##Does bb.nm line up with coords in the first place? YES
plot(bb.nm$geometry)
plot(c2$geometry, add=T, col="maroon", alpha=.05)

# view!
library(rnaturalearth)
library(rnaturalearthdata)
library(rgeos)
world <- ne_countries(scale = "medium", returnclass = "sf")

###To check that sf layers are aligning:::
plot(bb.nm$geometry, pch=16, cex=0.2, col="white")
#plot(c2$taxa, col="goldenrod", pch=16)
plot(c2$geometry, add=T, col="maroon", cex=.1, pch=18, alpha=.02)
title("Herbarium collections across NM counties")

#First spatial join between dat_cl and the county data. 
#note that object `env.v` will be the working object that I'll add things to as I work in this script. Eventually save it as a new version of dat_cl
env.v <- st_join(c2, left = TRUE, bb.nm["COUNTYFP"]) # join points IT WORKED!!!!!!! but somehow gained 265 records...
class(env.v)

check <- st_difference(c2, bb.nm["COUNTYFP"])

###PLOT county collections
ggplot(bb.nm)+
  geom_sf(aes(fill=COUNTYFP))+
  theme(legend.position = "bottom")

bbnm1<-env.v %>% 
  group_by(COUNTYFP) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  mutate_at("count", as.numeric)
bbnm1

ggplot(bbnm1, aes(x=reorder(COUNTYFP, -count), y=count, fill=COUNTYFP))+
  geom_bar(stat="identity") +
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

##Environmental data
There are multiple ways to access environmental data useful for geospatial analyses. We will use environmental data loaded into R using packages with API connections, as well as raster data downloaded from open source geospatial data providers. There are often high resolution or tailored geospatial data products developed for specific uses or regions; these can be quite pricey, or may require institutional affiliation to gain access. Our objective here is to make scripts fully reproducible, so we do not include any layers that require payment or institutional privileges. Users may want to consider additional sources for their own analyses, however. Rasters from many sources can be added into the raster stack (brick?) created in ADD CHUNK

NOTE THAT YOU MAY HAVE TO SWITCH OFF SPHERICAL GEOMETRY (S2) USING THE FOLLOWING CODE:
sf::sf_use_s2(FALSE)


###SHAPEFILES
Load New Mexico shapefiles 

###1. Physiographic map of the usa that's not especially high res (not working)

```{r}
#switch off spherical geometry
sf::sf_use_s2(FALSE)

#physiographic map
physio <- st_read("/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Environmental Data/Layers to be used/physio_shp/physio.shp", crs=4326)
physio$physio.area <- st_area(physio) #calculate area from geometries
plot(st_geometry(physio)) #this is the physiographic map of the entire USA

#Crop the full Physio layer to NM only
#crop<-st_crop(physio, xmin=-109.05, ymin=31.33224, xmax=-103.0022, ymax=37.00014)
#nmphysio2<-st_crop(physio.geo, st_bbox(geometry)) #I would rather crop to the NM wkt/boundary, but it's not working..

#Are the coordinates aligning with map layer? YEP
plot(physio$geometry, col=alpha("lightgray", 0.5), border="black", lwd=1.5)
plot(aoi, add=T, color="darkgray", alpha=.2)
plot(c2$geometry, add=T, col="maroon", pch=16, cex=.02)
title("Physiographic map of USA with NM coordinates")


#Join physio and coords to extract values
sf::sf_use_s2(FALSE)
env.v <- st_join(c2, left = TRUE, physio["PROVINCE"]) %>% st_join(left=TRUE, physio["physio.area"])

                                                                  
#env.v <- env.v %>% st_join(left = TRUE, physio["PROVINCE"]) %>% st_join(left = TRUE, physio["physio.area"]) #create new 


###BASIC PLOTS
ggplot(physio)+
  geom_sf(aes(fill=PROVINCE))+
  geom_sf(data=aoi, aes(alpha=0.02))+
  theme(legend.position = "bottom")

phys1<-env.v %>% group_by(physio.reg) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  mutate_at("count", as.numeric)
phys1

ggplot(phys1)+
  geom_sf(aes(col=physio.reg))+
  theme(legend.position = "right")

ggplot(phys1, aes(x=reorder(physio.reg, -count), y=count, fill=physio.reg))+
  geom_bar(stat="identity") +
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

#save
ggsave(filename = "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Figures/phys1.png", width = 16, height = 10)

```

### 2. Geologic Regions
```{r}
#geologic map
georeg <- st_read("/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Environmental Data/Layers to be used/slo_GeologicRegions/slo_oggeolregions.shp")
georeg <-st_transform(georeg, crs=4326) #necessary step to get coords in the right alignment
georeg$georeg.area <- st_area(georeg) #calculate area from geometries
plot(st_geometry(georeg)) #this is the geologic map of NM

#Are the coordinates aligning with map layer? 
plot(georeg$geometry, col=alpha("lightgray", 0.5), border="black", lwd=1.5)
plot(c2$geometry, add=T, col="aquamarine4", pch=16, cex=.1, alpha=.02)
title("Geologic map of USA with NM coordinates")

#Join geologic and coords to extract values
env.v <- env.v %>% st_join(left = TRUE, georeg["LABEL"])  %>% st_join(left = TRUE, georeg["georeg.area"])



###BASIC PLOTS
ggplot(georeg)+
  geom_sf(aes(fill=LABEL))+
  theme(legend.position = "r")

g1<-env.v%>%group_by(geologic) %>%
  dplyr::filter(!geologic=="NA") %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  mutate_at("count", as.numeric)
g1

ggplot(g1)+
  geom_sf(aes(col=geologic))+
  theme(legend.position = "right")

ggplot(g1, aes(x=reorder(geologic, -count), y=count, fill=geologic))+
  geom_bar(stat="identity") +
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

#save
ggsave(filename = "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Figures/phys1.png", width = 16, height = 10)

```

###3. Ecoregions: still need to add to the env.v dataframe
```{r}
#ecoregion map
ecoreg.l4 <- st_read("/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Environmental Data/Layers to be used/nm_ecoregions_l4/nm_eco_l4.shp")
 ecoreg.l4 <-st_transform(ecoreg.l4, crs=4326) #necessary step to get coords in the right alignment
ecoreg.l4$ecoreg.area <- st_area(ecoreg.l4) #calculate area from geometries

plot(st_geometry(ecoreg.l4)) #this is the ecoregion map of NM

#Are the coordinates aligning with map layer? YEP
plot(ecoreg.l4$geometry, col=alpha("lightgray", 0.5), border="black", lwd=1.5)
plot(c2$geometry, add=T, col="tomato3", pch=16, cex=.1, alpha=.5)
title("Ecoregion map of USA with NM coordinates")

#Join ecoregion and coords to extract values
env.v <- env.v %>% st_join(left = TRUE, ecoreg.l4["US_L4NAME"])  %>% 
  st_join(left =TRUE, ecoreg.l4["US_L3NAME"]) %>%
  st_join(left = TRUE, ecoreg.l4["ecoreg.area"]) 



#Try plotting to see about biases
eco4<-env.v%>%group_by(ecoreg4) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  mutate_at("count", as.numeric)

eco3<-env.v%>%group_by(ecoreg3)%>%
  summarise(count = n()) %>%
  arrange(desc(count))%>%
  mutate_at("count", as.numeric)


#Quick look at the distribution of collections from different ecoregions
ggplot(eco4, aes(x=reorder(ecoreg4, -count), y=count, col=ecoreg4))+
  geom_bar(stat="identity", fill="gray34") +
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggplot(datgeo, aes(ecoreg4, ecoreg.area)) +
      geom_bar(stat="identity", col="black")

ggplot(eco3, aes(x=reorder(US_L3NAME, -count), y=count, fill=ecoreg3)) +
  geom_bar(stat="identity") +
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle =90, vjust = 1, hjust=1))

#save
ggsave(filename = "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/eco4.png", width = 16, height = 10)

ggsave(filename = "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Figures/eco3.png", width = 16, height = 10)


##Plot number of taxa (count) divided by the area of each ecoregion


eco3<-st_join(eco3, area.eco3)      


ggplot(eco3, aes(x=reorder(US_L3NAME.x, -count), y=count, fill=as.factor(area.sum)))+
  geom_bar(stat="identity")+
  theme(legend.position="none")+
  theme(axis.text.x = element_text(angle =90, vjust = 1, hjust=1))

ggplot(eco3, aes(count, area.sum, col=US_L3NAME.x, label=US_L3NAME.x))+
  geom_jitter(cex=3)+
  geom_text(hjust=.5, vjust=-1)+
  theme(legend.position="none")

ggsave(filename = "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Figures/eco3areacount.png", width = 16, height = 10)


####Replicate for Ecoregions layer 4
##Plot number of taxa (count) divided by the area of each ecoregion
area.eco4<-ecoreg.l4 %>% group_by(US_L4NAME) %>% 
  summarise(area.sum=sum(Shape_Area/1000))

eco4<-st_join(eco4, area.eco4)      

ggplot(eco4, aes(area.sum, count, col=US_L4NAME.x))+
  geom_jitter(cex=2)+
  theme(legend.position="none")+
  geom_text(aes(label=ifelse(area.sum>7000000,as.character(US_L4NAME.x),'')),hjust=.5,vjust=-1, cex=4)+
  geom_text(aes(label=ifelse(count>8000,as.character(US_L4NAME.x),'')),hjust=.5,vjust=-1, cex=4)


ggsave(filename = "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Figures/eco4areacount.png", width = 16, height = 10)


##Rather than just total number of collections, check out the proportion of all collections
ggplot(eco4, aes(area.sum, count/sum(count), col=US_L4NAME.x))+
  geom_jitter(cex=2)+
  theme(legend.position="none")+
  geom_text(aes(label=ifelse(area.sum>7000000,as.character(US_L4NAME.x),'')),hjust=.5,vjust=-1, cex=5)+
  geom_text(aes(label=ifelse(count>8000,as.character(US_L4NAME.x),'')),hjust=.5,vjust=-1, cex=5)
ggsave(filename = "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Figures/eco4areacount.v2.png", width = 16, height = 10)


###TRY plotting the map of ecoregions with colors
ggplot(ecoreg.l4)+
  geom_sf(aes(fill=US_L3NAME))+
  theme(legend.position = "bottom")

ggsave(filename = "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Figures/eco3map.png", width = 16, height = 10)


ggplot(ecoreg.l4)+
  geom_sf(aes(fill=US_L4CODE))+
  geom_sf(data=c2, aes(x=long, y=lat), cex=.1, alpha=.3)+
  theme(legend.position = "none")

ggsave(filename = "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Figures/eco4map+coords.png", width = 16, height = 10)

```

### 4. Land ownership

```{r}
#land ownership shapefile map
owner <- st_read("/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Environmental Data/Layers to be used/SURFACE_OWN/Surface_Ownership.shp")
owner <-st_transform(owner, crs=4326) #necessary step to get coords in the right alignment
owner$own.area <- st_area(owner) #calculate area of land parcels

#The shapefile is a bit messed up. https://stackoverflow.com/questions/68478179/how-to-resolve-spherical-geometry-failures-when-joining-spatial-data/
sf::sf_use_s2(FALSE)

plot(st_geometry(owner)) #this is the land ownership map of NM

#Are the coordinates aligning with map layer? YES
plot(owner$geometry, col=alpha("lightgrey", 0.5), lwd=.05)
plot(c2$geometry, add=T, col="darkblue", pch=16, cex=.05, alpha=.5)
title("Herbarium collections across land ownership categories in NM")

#Join ownership areas and coords to extract values
env.v <- env.v %>% st_join(left = TRUE, owner["own"])  %>% st_join(left = TRUE, owner["own.area"])


#The shapefile is a bit messed up. Turn s2 back on for the newest version of sf 
sf::sf_use_s2(TRUE)


###Native land numbers
natland<- env.v %>%
  filter(ownership=="I")

```


### 5. ****TNC Southwest biotic communities

```{r}
#TNC sw biotic map
swbio <- st_read("/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Environmental Data/Layers to be used/TNC SW biotic communities/sw_biotic_communities.shp")
swbio <-st_transform(swbio, crs=4326) #necessary step to get coords in the right alignment
swbio.geo <- st_geometry(swbio)
swbio$swbio.area <- st_area(swbio) #calculate area of swbio regions

plot(st_geometry(swbio)) #this is the geologic map of NM

#Are the coordinates aligning with map layer? 
plot(swbio$geometry, col=alpha("white", .5), border="black", lwd=.5)
plot(c2$geometry, add=T, col=alpha("cornflowerblue", 0.5), pch=16, cex=.1, alpha=.5)
title("TNC southwest biotic communities with NM coords")

#Join geologic and coords to extract values
env.v <- env.v %>% st_join(left = TRUE, swbio["COMMUNITY"])  %>% st_join(left = TRUE, swbio["swbio.area"])


###Basic Plotting
swb1<-env.v%>%group_by(swbiocomm) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  mutate_at("count", as.numeric)
head(swb1)

ggplot(swb1, aes(x=reorder(swbiocomm, -count), y=count, fill=swbiocomm))+
  geom_bar(stat="identity") +
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```


### 6. Conservation and critical concern areas (MAYBE join multiple layers? Or choose one?)
```{r}
acec<- st_read("/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Environmental Data/Layers to be used/ACEC/ACEC.shp") %>% st_transform(crs=4326)
blm<-st_read("/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Environmental Data/Layers to be used/BLM_NM_WILDERNESS_BOUNDARIES/BLM_NM_WILDERNESS_BOUNDARIES.shp") %>% st_transform(crs=4326)
nca<-st_read("/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Environmental Data/Layers to be used/NLCS_NM_NCA/NLCS_Natural_Conservation_Areas.shp") %>% st_transform(crs=4326)
study<-st_read("/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Environmental Data/Layers to be used/NLCS_WILDERNESS_STUDY_AREAS/NLCS_Wilderness_Study_Areas.shp") %>% st_transform(crs=4326)

#add column of protected lands area 
acec$acec.area <- st_area(acec) #calculate area from geometries

#First try plotting multiple layers at once to see how they line up. Pretty similar. 
plot(st_geometry(aoi), col="lightgrey")
plot(st_geometry(acec), col="cornflowerblue", add=TRUE)
plot(st_geometry(nca), col="darkred", alpha=.2, add=TRUE)
plot(st_geometry(study), col="orange", add=TRUE)
plot(st_geometry(blm), col="gold", add=TRUE)



#Keep individual values in each layer, but also create a new multipolygon layer that is the intersection of all four 'protected' land types
p.multi<- st_union(acec, blm, nca, study, by_feature=TRUE) 
plot(st_geometry(p.multi), col="purple") #I'm not sure this is actually right but I'm going to give this a go. 

# try2<-st_difference(acec,st_union(st_combine(blm)))
# plot(st_geometry(try2))

#Are the coordinates aligning with map layer? 
plot(p.multi$geometry, col=alpha("darkgray", .5), border="black", lwd=.5)
plot(c2$geometry, add=T, col=alpha("purple4", 0.05), pch=16, cex=.2, alpha=.5)
title("Map of concatenated protected land in NM")

#Join protected areas and coords to extract values
#env.v <- st_join(env.v, left = TRUE, acec["ACEC_NAME"])  #need to figure out a different way of creating a 'protected' or 'unprotected' column. If else statement probably
env.v <- env.v %>% st_join(left = TRUE, acec["ACEC_NAME"])  %>% st_join(left = TRUE, acec["acec.area"])



###BASIC PLOTS
ggplot(p.multi)+
  geom_sf(aes(fill=ACEC_NAME))+
  geom_sf(data=aoi, aes(alpha=0.2))+
  theme(legend.position = "r")

protect1<-env.v%>%group_by(protection) %>%
  dplyr::filter(!protection=="NA") %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  mutate_at("count", as.numeric)
protect1

ggplot(protect1)+
  geom_sf(aes(col=protection))+
  theme(legend.position = "r")

ggplot(protect1, aes(x=reorder(protection, -count), y=count, fill=protection))+
  geom_bar(stat="identity") +
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, size=4, vjust = 0.5, hjust=1))

#save
ggsave(filename = "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Figures/protectbarplot.png", width = 16, height = 10)


```

###Save df of env.v data (shapefiles)
There will also be a raster version of this, I think. The key df here is 'env.v', which is a part of our original cleaned df (nmcol_resolved), then converted to sf format and joined with a bunch of environmental data in columns. The environmental data here is from vector layers (shapefiles). We will join this df with raster environmental data next. 
```{r}
env.v<-env.v %>% rename("countyFP" = "COUNTYFP",
                         "physio" = "PROVINCE",
                         "geologic" = "LABEL",
                         "ecoreg4" = "US_L4NAME",
                         "ecoreg3"= "US_L3NAME",
                         "ownership" = "own",
                         "swbiocomm" = "COMMUNITY",
                         "protection" = "ACEC_NAME")

class(env.v) #sf data frame 

##save rda
save(env.v, file="GBIFgeoDF_withenvironmental.RData")

write.csv(env.v, "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Environmental Data/environ_Dat_cl_Georef.csv")

env.v<-read.csv("/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Environmental Data/environ_Dat_cl.csv", header=TRUE)

```


##RASTER LAYERS

Land Use/Land Cover data
To get this layer, researchers should go to the MRLC website (https://www.mrlc.gov/viewer/) and download data for their region of interest. This may take up to a day to process. You will recieve a zipped file with raster(s). 



###Climate data rasters
For now I'll use prism to download PRISM data. This might not work... but I'm going to give in another go. Also...lower res might be good given the likelihood of bias in our occurrence data, too?

```{r}
library(prism)

#Specify where raster layers will go
options(prism.path="/Users/elizabethlombardi/Desktop/PRISM")
#Set folder for where PRISM layer downloads go
prism_set_dl_dir("/Users/elizabethlombardi/Desktop/PRISM") 

#Download normals data for mean temperature in July
#get_prism_normals(type = 'tmin', resolution = '4km', keepZip = TRUE)
#precip<-get_prism_normals("ppt", resolution="4km", annual=TRUE)



######CHOOSE WHICH PRISM DATA TO USE AND RE-RUN
#unlink(paste0("/Users/elizabethlombardi/Dropbox/PRISM"), recursive = TRUE)  #if you need to remove the files 

# Download the 30-year annual average precip and annual average temperature
get_prism_normals("ppt", "4km", annual = TRUE, keepZip = FALSE)
get_prism_normals("tmean", "4km", annual = TRUE, keepZip = FALSE)

#Try annual PRISM averages from 2010 to 2022. 4km resolution
get_prism_annual("ppt", years = 2010:2022, keepZip = FALSE)
get_prism_annual("tmean", years = 2010:2022, keepZip = FALSE)
######



#check out the list of rasters I now have
prism_archive_ls()
prism_get_dl_dir()


# Grab the prism data and compile the files into a raster stack (precip and temp)
climate_data <- prism_archive_ls() %>%  
  pd_stack(.)  

# Extract project coordinates from raster stack
climate_crs <- climate_data@crs@projargs


# Convert collection locations to format and projection that matches prism data
# make spatial
env.v <- env.v %>% 
  mutate_at(c("decimalLongitude", "decimalLatitude", "id"), as.numeric) %>%
  st_as_sf(coords=c("decimalLongitude","decimalLatitude"), crs=CRS(climate_crs), remove=FALSE) %>%
  as.data.frame(.) 

coordinates(env.v) <- c('decimalLongitude', 'decimalLatitude')
proj4string(env.v) <- CRS(climate_crs)


# Extract the  extracted from the raster stack for those sites 
prism.df <- data.frame(coordinates(env.v), 
                  env.v$id,
                  extract(climate_data, env.v))



#rename columns with annual mean precip and temp values for each row
prism.df <- prism.df %>% #includes provisional 2022 data
#  mutate_at(c("env.coords.id"), as.numeric) %>%
  rename("precip30yr" = "PRISM_ppt_30yr_normal_4kmM4_annual_bil") %>% 
  rename("temp30yr" = "PRISM_tmean_30yr_normal_4kmM4_annual_bil") %>%
  rename("prismLong" = "decimalLongitude") %>%
  rename("prismLat" = "decimalLatitude")
  

#convert env.v back to dataframe
env.v <- as.data.frame(env.v)


# #keep them in simple feature format? NOPE this doesn't seem to work. It might...but it takes FOREVER and then breaks. 
# try<- st_as_sf(env.v)
# try2<-prism.df %>% st_as_sf(coords=c("prismLong", "prismLat"), crs=crs(env.v))
# join <- try %>% st_join(left = TRUE, try2["precip30yr"])  %>% st_join(left = TRUE, try2["temp30yr"])


#join prism.means with env.v using id columns
env.v <- left_join(env.v, prism.df, by=(c("id"="env.v.id"))) #this joins PRISM 30-year normals to env.v dataframe, and also two new lat/long columns. Useful to make sure the abiotic data line up with the right collections. 

all(env.v$decimalLatitude.x == env.v$decimalLatitude.y) #again, just make sure the new lat/long coords match the old ones, which confirms if the join worked
dim(env.v) #243890    103


#save new env.v 
write.csv(env.v, "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Environmental Data/environ_Dat_cl.csv")


##save rda
save(env.v, file="georefData_PRISM.RData")


```


###Elevation data rasters
```{r}
library(elevatr)

load(file="script4_vectorPrism_Data.RData")

prj_dd <- "EPSG:4326"

df<-data.frame(env.v) %>%  #subsetting
  dplyr::select(x="decimalLongitude", y="decimalLatitude", id="id") %>% 
  mutate_if(is.character, as.numeric)

df.sub <- df %>% slice(1:100) #subset to look at first 100 records

elev<- get_elev_point(df, prj=prj_dd, src="epqs") #this step takes a pretty long time!!

elev<-as.data.frame(elev)
  


##Try getting a NM raster of elevation data instead
library(rgeoboundaries)
library(sf)
library(raster)
library(ggplot2)
library(viridis)

library(rnaturalearth)
library(rnaturalearthhires)

nm_map <- get_stamenmap(
  bbox = c(left = -109.5, bottom = 31, right = -102.8, top = 37.5), 
  maptype = "toner-lite",
  zoom = 7
)



swiss_bound <- rgeoboundaries::geoboundaries("Switzerland")
elevation_data <- elevatr::get_elev_raster(locations = swiss_bound, z = 9, clip = "locations")


elevation_data <- as.data.frame(elev, xy = TRUE)
colnames(elevation_data)[3] <- "elevation"
# remove rows of data frame with one or more NA's,using complete.cases
elevation_data <- elevation_data[complete.cases(elevation_data), ]

ggplot() +
  geom_raster(data = elevation_data, aes(x = x, y = y, fill = elevation)) +
  #geom_sf(data = df.sub, color = "white", fill = NA) +
  coord_sf() +
  scale_fill_viridis_c() +
  labs(title = "Elevation", x = "Longitude", y = "Latitude", fill = "Elevation (meters)")

```



##NM shaded relief 
Downloaded from the internet (open source) and uploading a .tif file here 

```{r}
library(raster)
library(terra)

nmSr <- raster::raster("/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Environmental Data/nm_relief_color_tif/nm_relief_color.tif")

nmSr <- terra::rast("/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Environmental Data/nm_relief_color_tif/nm_relief_color.tif")
nmSr <- project(nmSr, prj_dd)
plot(nmSr)
nlyr(nmSr) #three layers in the file
res(nmSr)

```

###Try again with a raster of NM download and then extract the data
```{r}
library(rasterVis)
library(raster)
library(rgl)
library(rgdal)
library(elevatr)


elevation <- get_elev_raster(aoi, z = 9)
plot(elevation, main="This the downloaded DEM [meters]")
plot(aoi$geometry, add=TRUE)

#crop to NM boundary
elevNM_crop = crop(elevation, aoi)
plot(elevNM_crop, main="Cropped Digital elevation model")
plot(aoi$geometry, add=TRUE)
plot()

elevNM_crop


#reproject
# Project Raster
# to have control,  we create one raster object
# using projectExtent (no values are transferred)
pr3 <- projectExtent(elevNM_crop, prj_dd)
# Adjust the cell size 
res(pr3) <- 100
# now project
rep_elev <- projectRaster(elevNM_crop, prj_dd)
hist(elevNM_crop)

# Grab the elevation data and compile the files into a raster stack
elev_data <- pd_stack(.)  

# Extract project coordinates from raster stack
elev_crs <- elevNM_crop@crs@projargs


# Convert collection locations to format and projection that matches prism data
# make spatial
env.v <- env.v %>% 
  mutate(rowID=row_number())%>%
  mutate_at(c("decimalLongitude", "decimalLatitude", "rowID"), as.numeric) %>%
  filter(!decimalLongitude=="", !decimalLatitude=="") %>%
  st_as_sf(coords=c("decimalLongitude","decimalLatitude"), crs=CRS(elev_crs), remove=FALSE) %>%
  as.data.frame(.) 
dim(env.v)


coordinates(env.v) <- c('decimalLongitude', 'decimalLatitude')
proj4string(env.v) <- CRS(elev_crs)


# Extract the  extracted from the raster stack for those sites 
elev.df <- data.frame(coordinates(env.v), 
                  env.v$rowID,
                  extract(elevNM_crop, env.v))

#rename columns 
elev.df <- elev.df %>% 
  rename("rowID" = "env.v.rowID") %>%
  rename("elevation" = "extract.elevNM_crop..env.v.") %>%
  rename("elevLat"="decimalLongitude")%>%
  rename("elevLong"="decimalLatitude")
  

#convert env.v back to dataframe
env.v <- as.data.frame(env.v)


#join elev.df with env.v using id columns
env.v <- left_join(env.v, elev.df, by="rowID") #this joins the elevation dataframe to the env.v main dataframe. Why did this increase the number of rows?

all(env.v$decimalLatitude.x == env.v$elevLat) #again, just make sure the new lat/long coords match the old ones, which confirms if the join worked
dim(env.v) 


#save new env.v 
write.csv(env.v, "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Full_Dataframe_georeffed.csv")


##save rda
save(env.v, file="georefEnviroFull.RData")



```





#Joining environmental dataframes
Specifically, I'm wondering if and how to join different layers with count information together. There isn't a unique ID in many of these layers, unless we can use geometry, for example? Assuming there aren't multiple collections that have exactly the same geometry/coords?

```{r}
nm.eco.df<-as.data.frame(nm.eco)
try<-st_join(nm.eco, c2) #Nope
```


