---
title: "Manuscript figures"
author: "emlombardi"
date: "2023-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Draft figures for the manuscript
you will likely have to read in the datafiles in question. nmcol_resolved if it's already loaded in R, or load MasterDF_spattax_cleaned.csv from local machine. This is the cleaned data (spatial and taxonomically to species level as of Jan 16)

#libraries
```{r}
library(ggplot2)
library(ggpubr)
library(khroma) #for Tol color palette stuff

theme_set(theme_pubr())
iridescent <- colour("iridescent")

#load
#save full working environment
dat <-load(file="script3secondtry_output.RData")
dat <- env.v
names(dat)

```

## Figure 1: Taxonomic trends
here we want to plot the distribution of collections for each species. On the y axis is the count of species that occur in each bin. On the x axis is the number of collections/species (1, 2, ...)

```{r pressure, echo=FALSE}
dat<-alldat

#species diversity
spp <- dat %>%
  group_by(Accepted_name) %>%
  filter(!Accepted_name == "") %>% #there are 2398 records without species epithet 
  dplyr::summarise(count = n())

hist(spp$count, breaks=100)
summary(spp$count)

top15<-dat %>% 
  add_count(Accepted_species) %>% 
  filter(n %in% tail(sort(unique(n)), 15)) %>% 
  filter(!Accepted_species == "") %>%
  arrange(desc(n)) #table of all records for the ten best represented species
top15list<-c(top15$Accepted_species)

ggplot(top15, aes(Accepted_species))+
  geom_bar() +
  theme_pubclean()

# Compute the frequency
library(dplyr)
fig1 <- dat %>%
  group_by(Accepted_name)%>%
  filter(!Accepted_name == "") %>%
  tally(sort=TRUE)
fig1


# Create the bar plot. Use theme_pubclean() [in ggpubr]
f1 <-ggplot(data = fig1, aes(n)) +
  geom_histogram(color = "darkslategrey", stat="count", linewidth=3) +
  scale_x_log10()+ #log the count
  labs(title = "How many collections do we have per species? (log transformation)")
f1
ggsave("/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Figures/logScale_noSpp.png", f1, width = 10, height = 5, dpi = 300)


singles<-fig1%>%filter(n=="1") #1285 species with a single collection

print(fig1)
summary(fig1)

#Try a boxplot to see things a bit differently. 
f1b<-ggplot(fig1, aes(x = factor(1), y = n)) +
  geom_boxplot(width = 0.7, fill = "white") +
  labs(x = NULL)   # Remove x axis label
f1b


#go up a taxonomic level
fig1c <- dat %>%filter(!Genus_matched=="")%>%
  group_by(Genus_matched)%>%tally(sort=TRUE)
fig1c

f1c<-ggplot(data = fig1c, aes(n)) +
  geom_bar(color = "coral4", stat="count", size=1) +
  scale_x_log10()+ #log the count
  labs(title = "How many collections do we have per genus? (log transformation)")
f1c

singles.genus<-fig1c%>%filter(n=="1") #240 genera with only a single collection

#go up a taxonomic level
figfam <- dat %>%filter(!family=="")%>%
  group_by(family)%>%tally(sort=TRUE)
figfam


###Look at only those species with more than one collection
fig1d<-fig1%>%filter(!n=="1") #Only those with 2+ species
f1d<-ggplot(data = fig1d, aes(n)) +
  geom_bar(color = "steelblue", stat="count", size=1) +
  labs(title = "How many species have multiple collections?")
f1d

```

##Donut plots
Using webr package to do this. Check out the tutorial here https://statdoe.com/pie-donut-chart-in-r/
```{r}
# loading the appropriate libraries
library(ggplot2)
library(webr)
library(dplyr)

# Building a table with the data for the plot
donut1 = dat %>%
  group_by(Taxonomic_status, Accepted_family) %>% 
  tally(sort=TRUE) %>% 
  filter(!Accepted_family=="", !Taxonomic_status=="Invalid") 

pd1<-PieDonut(donut1, aes(Taxonomic_status, Accepted_family, count=n), title = "Taxonomic status: count of families", explode=2, r0=.6)

pd1

#look more closely at the synonyms
donut2 = dat %>%  #21,279 of all records were taxonomic synonyms
  filter(Taxonomic_status=="Synonym") %>%
  filter(!Accepted_family=="", !Taxonomic_status=="Invalid") %>%
  group_by(Accepted_family, Genus_submitted) %>% 
  tally(sort=TRUE)

pd2<-PieDonut(donut2, aes(Accepted_family, Genus_submitted, count=n), title = "Taxonomic status: count of synonymous families", explode=2, r0=.6, color="white")
pd2 #this is super messy. The table donut2 is much more useful. 


# Create an interactive donut chart with Plotly!
pie1<-plot_ly(donut2, labels = ~Accepted_family, values = ~n, type = 'pie',
             hole = 0.45, textposition = "inside",
             textinfo = "label+percent",
             marker = list(colors = c("#FFA07A", "#87CEFA", "#32CD32", "#DA70D6", "#FFD700"),
                           line = list(color = "#FFFFFF", width = .25)))
pie1

ggsave("/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Figures/familyPiePlot.png", pie1, width = 10, height = 5, dpi = 300)

plotly_IMAGE(pie1, format = "png", out_file = "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Figures/familyPiePlot.png")

        
```

###Donut from Jansen's code
Try using code from Jans to create donut plots in ggplot format
```{r}
## donut plot for proportion contribution of each database
don3 <- dat_cl %>%
  group_by(source) %>%
  dplyr::summarise(count = n())%>%
  filter(!count<100)

# Compute percentages
don3$frac <- don3$count / sum(don3$count)

# Compute the cumulative percentages (top of each rectangle)
don3$ymax <- cumsum(don3$frac)

# Compute the bottom of each rectangle
don3$ymin <- c(0, head(don3$ymax, n=-1))

# Compute label position
don3$labelPosition <- (don3$ymax + don3$ymin)/2

# Compute a good label
don3$label <- paste0(don3$order, ", ", don3$count)

# Make the plot
png(file.path("export",ver,"bdt_dbDonut.png"), width=2000, height=2000)

ggplot(don3, aes(ymax=ymax, ymin=ymin, xmax=4, xmin=3, fill=source)) +
	geom_rect() +
	#geom_text( x=2, aes(y=labelPosition, label=label), size=6) + # x here controls label position (inner / outer)
	scale_color_viridis("magma") +
	coord_polar(theta="y") +
	xlim(c(2, 4)) +
	theme_void() #+
	#theme(legend.position = "none")
dev.off()


```


#Figure 2: Temporal trends
Are there trends in when collections have been made? By year and by month/julien date?

Idea for better figure/analysis: Find the right resource to determine monsoon seasonality and look at temporal trends in collections based on historical monsoons? Not reproducible unless I can find a publically-available rain resource (I can) that would be generalizable...

```{r temporal}

#fig2a: calculate the percent of total collections made each year

fig2<-dat %>%group_by(year)%>%tally(sort=TRUE)%>%na.omit()


time <- count(dat, year, Accepted_species, order)  %>% group_by(n)

f2a<-ggplot(data = time, aes(x = year, y = n)) +
    geom_line()+
  labs(title="Collected species by year since 1800")
f2a



###
time2<-dat %>% group_by(year)%>% filter(!year<1800)%>%tally(sort=TRUE) %>% na.omit() %>% mutate(ratio=n/sum(n))

f2b<-ggplot(data = time2, aes(year, ratio)) +
  geom_point(color = "slategray4",size = 2) +
  geom_line(color="slategray4", stat="") + 
  labs(title = "Percent of total holdings made each year")
f2b


###
time3 <- time %>% group_by(year) %>% mutate(spp=n_distinct(Accepted_species))


f2c<-ggplot(data = time3, aes(year, spp)) +
  geom_point(color = "turquoise4", size =1) +
  labs(title = "Number of species collected each year")
f2c



####Figure 2 temporal stats and trends in seasonality
#https://cran.r-project.org/web/packages/funtimes/vignettes/trendtests.html
#fig2a: calculate the percent of total collections made each year

fig2d<-dat%>%group_by(date.ordinal)%>%tally(sort=TRUE)%>%na.omit()


time <- count(dat, date.ordinal, taxized)  %>% group_by(n)
f2d<-ggplot(data = time, aes(x = date.ordinal, y = n)) +
    geom_line()+
  labs(title="Collected species over ordinal time")
f2d



####Take out the taxonomic grouping and see how all 
time <-count(dat, date.ordinal) %>% group_by(n) %>% na.omit()
sp.yr<- ggplot(data = time, aes(x = date.ordinal, y = n)) +
    geom_line()+
    labs(title="Collected accessions over ordinal time")

###OTHERS


```
#Environmental trends
To get the next chunks to run, you need to have run the script '4. Download enviro data.Rmd', which is where we create spatially joined dataframes of occurrences and environmental variables. Load the 'env.v' csv file that is produced from the shapefiles (vector formatted environmental variables). 

Here I'll produce figures to i) count total collections by environmental variable categories

```{r}
#load data
env.v <- read.csv("/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Environmental Data/environ_shps.csv")

view(env.v)

try<- env.v %>% group_by(physio.area) %>% mutate(physio.Asum = sum(physio.reg))


ggplot(env.v, aes(x=physio.area, y=count(physio.reg)))+
  geom_bar()

##Plot number of taxa (count) divided by the area of each ecoregion
#Try plotting to see about biases
eco4<-env.v%>%group_by(ecoreg4) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  mutate_at("count", as.numeric)

area.eco4<-env.v %>% group_by(ecoreg4) %>% 
  summarise(area.sum=sum(ecoreg.area/1000)) %>%
  na.omit()

colnames(area.eco4)[2] <- gsub("\\s*\\[.*\\]", "", colnames(area.eco4)[2])


eco4<-st_join(eco4, area.eco4)    

eco4norm<- eco4 %>%
  mutate(norm.count = (count - min(count)) / (max(count) - min(count))) %>%
  mutate(ratio = (count/area.sum)*1000)  #change to kilometer 

eco4norm <- eco4norm %>%
  mutate(ratio.quants = quantile(ratio, probs = c(0, 0.25, 0.5, 0.75, 1)))

ggplot(eco4norm, aes(area.sum, norm.count, col=ratio))+
  geom_jitter(cex=2)+
  theme(legend.position="none")


###Ecoregion 3

eco3<-env.v%>%group_by(ecoreg3) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  mutate_at("count", as.numeric)

area.eco3<-env.v %>% group_by(ecoreg3) %>% 
  summarise(area.sum=sum(ecoreg.area/1000)) %>%
  na.omit()

eco3<-st_join(eco3, area.eco3)    
eco3norm<- eco3 %>%
  mutate(norm.count = (count - min(count)) / (max(count) - min(count))) %>%
  mutate(ratio = (count/area.sum)*1000) %>% #change to kilometer 
  filter(!ecoreg3.x == "")
  
ggplot(eco3norm, aes(area.sum, norm.count, col=ecoreg3.x))+
  geom_jitter(cex=4)+
  #geom_text(aes(label = ecoreg3.x), vjust = 2) +
  scale_fill_iridescent(reverse = FALSE) +
  ggplot2::theme_bw() +
  theme(legend.position="right")





ggsave(filename = "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Figures/eco4areacount.png", width = 16, height = 10)


##Rather than just total number of collections, check out the proportion of all collections


ggplot(eco4, aes(area.sum, count/sum(count), col=ecoreg4.x))+
  geom_jitter(cex=5)+
  theme(legend.position="none") +
  ggplot2::theme_bw() +
  scale_colour_bright()




ggsave(filename = "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Figures/eco4areacount.v2.png", width = 16, height = 10)


```

```{r}
inst <- dat %>%
  group_by(institutionCode) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

```



###TRY plotting the map of ecoregions with colors
ggplot(ecoreg.l4)+
  geom_sf(aes(fill=US_L3NAME))+
  theme(legend.position = "bottom")

ggsave(filename = "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Figures/eco3map.png", width = 16, height = 10)


ggplot(ecoreg.l4)+
  geom_sf(aes(fill=US_L4CODE))+
  geom_sf(data=c2, aes(x=long, y=lat), cex=.1, alpha=.3)+
  theme(legend.position = "none")

ggsave(filename = "/Users/elizabethlombardi/Desktop/Research/UNM/Herbarium collections patterns and gaps project/Figures/eco4map+coords.png", width = 16, height = 10)



```
